---
title: Comparing Eye-Movement Patterns
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Eye-Movement Patterns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-violet');});</script>

---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE))
  ggplot2::theme_set(albersdown::theme_albers(params$family))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

```{r load-packages}
library(eyesim)
library(dplyr)
library(ggplot2)
library(patchwork)
```

When you study an image, you move your eyes to different locations in a
particular sequence. If you later see that same image during a memory test, do
you look at the same places? This kind of "eye-movement reinstatement" is a
powerful marker of memory, and `eyesim` gives you the tools to measure it.

This vignette walks you through the core workflow: representing fixations,
computing density maps, and measuring similarity between fixation patterns
across experimental conditions.

## How do you represent fixations?

A `fixation_group` holds a set of eye fixations — each defined by an x/y
screen position, an onset time (when the fixation began), and a duration (how
long the eye stayed there). You can create one directly:

```{r create-fixation-group}
fg <- fixation_group(
  x = c(-100, 0, 100),
  y = c(0, 100, 0),
  onset = c(0, 10, 60),
  duration = c(10, 50, 100)
)
```

```{r plot-fixation-group, echo = FALSE, fig.cap = "Three fixations. Point size reflects duration; color reflects onset time (yellow = early, red = late)."}
plot(fg)
```

Point size shows how long each fixation lasted. Color indicates when it
occurred: yellow for early fixations, red for later ones.

Here is a more realistic group with 25 randomly placed fixations:

```{r random-fixations}
set.seed(42)
fg <- fixation_group(
  x = runif(25, 0, 100),
  y = runif(25, 0, 100),
  onset = cumsum(runif(25, 0, 100)),
  duration = runif(25, 50, 300)
)
```

```{r plot-random-fixations, echo = FALSE, fig.cap = "25 randomly placed fixations."}
plot(fg)
```

## How do you visualize fixation density?

Individual points can be hard to interpret. Density maps show you where
fixations cluster. The `plot()` method supports several display styles:

```{r density-types, fig.width = 9, fig.cap = "Three density visualizations: contour, raster, and filled contour."}
p1 <- plot(fg, typ = "contour", xlim = c(-10, 110), ylim = c(-10, 110), bandwidth = 35)
p2 <- plot(fg, typ = "raster", xlim = c(-10, 110), ylim = c(-10, 110), bandwidth = 35)
p3 <- plot(fg, typ = "filled_contour", xlim = c(-10, 110), ylim = c(-10, 110), bandwidth = 35)
p1 + p2 + p3
```

The `bandwidth` parameter controls the smoothing level. Higher values blur
out fine detail and emphasize broad patterns:

```{r density-bandwidth, fig.width = 9, fig.cap = "Effect of bandwidth: narrow (20), medium (60), and wide (100)."}
p1 <- plot(fg, typ = "filled_contour", xlim = c(-10, 110), ylim = c(-10, 110), bandwidth = 20)
p2 <- plot(fg, typ = "filled_contour", xlim = c(-10, 110), ylim = c(-10, 110), bandwidth = 60)
p3 <- plot(fg, typ = "filled_contour", xlim = c(-10, 110), ylim = c(-10, 110), bandwidth = 100)
p1 + p2 + p3
```

## How do you compare two fixation patterns?

To quantify how similar two fixation patterns are, convert them to
`eye_density` maps and call `similarity()`. Here we create two patterns that
share roughly half their fixation locations:

```{r create-pair}
set.seed(123)
x_shared <- runif(12, 0, 100)
y_shared <- runif(12, 0, 100)

fg1 <- fixation_group(
  x = c(x_shared, runif(13, 0, 100)),
  y = c(y_shared, runif(13, 0, 100)),
  onset = cumsum(runif(25, 0, 100)),
  duration = runif(25, 50, 300)
)

fg2 <- fixation_group(
  x = c(x_shared, runif(13, 0, 100)),
  y = c(y_shared, runif(13, 0, 100)),
  onset = cumsum(runif(25, 0, 100)),
  duration = runif(25, 50, 300)
)
```

```{r plot-pair, fig.width = 8, echo = FALSE, fig.cap = "Two fixation patterns sharing roughly half their locations."}
plot(fg1) + plot(fg2)
```

Now convert to density maps and compute their similarity:

```{r compute-similarity}
ed1 <- eye_density(fg1, sigma = 50, xbounds = c(0, 100), ybounds = c(0, 100))
ed2 <- eye_density(fg2, sigma = 50, xbounds = c(0, 100), ybounds = c(0, 100))
similarity(ed1, ed2)
```

The default metric is the Pearson correlation. Several alternatives are
available:

```{r similarity-methods}
methods <- c("pearson", "spearman", "fisherz", "cosine", "l1", "jaccard", "dcov")
results <- sapply(methods, function(m) similarity(ed1, ed2, method = m))
data.frame(method = methods, similarity = round(unlist(results), 4))
```

## How do you analyze a full experiment?

In a typical memory study, participants view images during encoding and again
during retrieval. You want to compare fixation patterns between these phases
for the same image and control for non-specific similarity.

Let's simulate a small experiment: 3 participants, 20 images, encoding +
retrieval.

```{r simulate-experiment, include = FALSE}
set.seed(99)

gen_fixations <- function(imname, phase, participant) {
  nfix <- sample(3:10, 1)
  data.frame(
    x = runif(nfix, 0, 100), y = runif(nfix, 0, 100),
    onset = cumsum(runif(nfix, 50, 200)),
    duration = runif(nfix, 100, 400),
    image = imname, phase = phase, participant = participant
  )
}

df <- do.call(rbind, lapply(c("s1", "s2", "s3"), function(s) {
  do.call(rbind, lapply(c("encoding", "retrieval"), function(p) {
    do.call(rbind, lapply(paste0("img", 1:20), function(img) {
      gen_fixations(img, p, s)
    }))
  }))
}))
```

```{r show-experiment}
head(df)
cat("Rows:", nrow(df),
    " | Participants:", length(unique(df$participant)),
    " | Images:", length(unique(df$image)))
```

Wrap the raw data in an `eye_table`, which groups fixations by the variables
that define your experimental design:

```{r create-eye-table}
eyetab <- eye_table("x", "y", "duration", "onset",
                    groupvar = c("participant", "phase", "image"),
                    data = df)
eyetab
```

### Computing density maps by group

`density_by()` computes a density map for every combination of your grouping
variables:

```{r compute-densities}
eyedens <- density_by(eyetab,
                      groups = c("phase", "image", "participant"),
                      sigma = 100,
                      xbounds = c(0, 100), ybounds = c(0, 100))
```

```{r plot-densities, fig.width = 8, echo = FALSE, fig.cap = "Four density maps from the simulated experiment."}
p1 <- plot(eyedens$density[[1]])
p2 <- plot(eyedens$density[[2]])
p3 <- plot(eyedens$density[[3]])
p4 <- plot(eyedens$density[[4]])
(p1 + p2) / (p3 + p4)
```

```{r show-densities}
eyedens
```

### Template similarity analysis

`template_similarity()` compares each retrieval density map to its matched
encoding density map. It estimates a baseline by permuting image labels and
reports the corrected difference:

```{r template-similarity}
set.seed(1234)
enc_dens <- eyedens %>% filter(phase == "encoding")
ret_dens <- eyedens %>% filter(phase == "retrieval")

simres <- template_similarity(enc_dens, ret_dens,
                              match_on = "image",
                              method = "fisherz",
                              permutations = 50)
```

The result includes three key columns:

- `eye_sim` — raw similarity between matched pairs
- `perm_sim` — average similarity across non-matching pairs (baseline)
- `eye_sim_diff` — the corrected score (raw minus baseline)

Since our data is random, there should be no true reinstatement:

```{r test-similarity}
t.test(simres$eye_sim_diff)
```

```{r plot-similarity, fig.width = 9, echo = FALSE, fig.cap = "Distribution of raw, permuted, and corrected similarity scores."}
library(tidyr)

sim_long <- simres %>%
  select(eye_sim, perm_sim, eye_sim_diff) %>%
  pivot_longer(everything(), names_to = "type", values_to = "similarity") %>%
  mutate(type = factor(type,
    levels = c("eye_sim", "perm_sim", "eye_sim_diff"),
    labels = c("Raw", "Permuted", "Corrected")))

ggplot(sim_long, aes(x = similarity)) +
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.7) +
  facet_wrap(~type, scales = "free_x") +
  labs(x = "Similarity (Fisher z)", y = "Count") +
  theme_minimal()
```

As expected, the corrected similarity is centered near zero.

**A note on units.** With `method = "fisherz"`, all similarity values are
Fisher z scores (atanh of Pearson r). Convert back to correlations with
`tanh()` if desired. With `method = "pearson"` or `"spearman"`, values are
correlations directly.

**A note on permutations.** The `permutations` argument is an upper bound. If
fewer non-matching candidates are available, `eyesim` uses all of them
exhaustively.

## What about multiscale analysis?

Choosing a single bandwidth is somewhat arbitrary. You can compute density at
multiple scales by passing a vector of sigma values:

```{r multiscale}
eyedens_multi <- density_by(eyetab,
                            groups = c("phase", "image", "participant"),
                            sigma = c(25, 50, 100),
                            xbounds = c(0, 100), ybounds = c(0, 100))

enc_multi <- eyedens_multi %>% filter(phase == "encoding")
ret_multi <- eyedens_multi %>% filter(phase == "retrieval")

simres_multi <- template_similarity(enc_multi, ret_multi,
                                    match_on = "image",
                                    method = "fisherz",
                                    permutations = 50)

cat("Single-scale mean:", round(mean(simres$eye_sim_diff), 4), "\n")
cat("Multiscale mean:  ", round(mean(simres_multi$eye_sim_diff), 4), "\n")
```

Multiscale analysis provides a more robust similarity estimate by averaging
across spatial resolutions.

## How does similarity evolve over time?

Sometimes you want to know *when* during retrieval participants look at
previously-fixated locations. `sample_density_time()` samples the encoding
density at retrieval fixation locations at each time point:

```{r temporal-similarity}
ret_eyetab <- eyetab %>% filter(phase == "retrieval")

enc_dens <- enc_dens %>%
  mutate(match_key = paste(participant, image, sep = "_"))
ret_matched <- ret_eyetab %>%
  mutate(match_key = paste(participant, image, sep = "_"))

temporal <- sample_density_time(
  template_tab = enc_dens,
  source_tab = ret_matched,
  match_on = "match_key",
  times = seq(0, 500, by = 50),
  time_bins = c(0, 250, 500),
  permutations = 10,
  permute_on = "participant"
)
```

```{r show-temporal}
temporal %>%
  select(participant, image, bin_1, bin_2,
         perm_bin_1, perm_bin_2, diff_bin_1, diff_bin_2) %>%
  head()
```

```{r plot-temporal, fig.width = 8, fig.height = 4, echo = FALSE, fig.cap = "Encoding density sampled at retrieval fixation locations over time, for six example trials."}
time_series <- temporal %>%
  head(6) %>%
  mutate(trial_id = paste(participant, image, sep = "_")) %>%
  select(trial_id, sampled) %>%
  tidyr::unnest(sampled)

ggplot(time_series, aes(x = time, y = z, color = trial_id)) +
  geom_line(linewidth = 1) +
  labs(x = "Time (ms)", y = "Sampled Density",
       title = "Encoding density at retrieval fixation locations") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

This reveals when during retrieval participants fixate locations that mattered
during encoding — a window into the time course of memory-guided attention.

## What's next?

- `vignette("Multimatch")` — compare scanpath structure across five dimensions
- `vignette("RepetitiveSimilarity")` — analyze repeated viewing of the same stimulus
- `vignette("latent-transforms")` — domain adaptation for cross-device or cross-participant comparisons
- The package ships real data: `data(wynn_study)` and `data(wynn_test)` from a recognition memory experiment
