---
title: "GazeWeave: Calibrated Item Information from Gaze Paths"
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
vignette: |
  %\VignetteIndexEntry{GazeWeave: Calibrated Item Information from Gaze Paths}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: lapis
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-violet');});</script>
---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE))
  ggplot2::theme_set(albersdown::theme_albers(params$family))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

```{r load}
library(eyesim)
library(dplyr)
library(ggplot2)
```

## The inferential object

GazeWeave asks whether a held-out gaze path identifies its study item.
It does not average several similarity dimensions. For each permitted candidate
template it fits the same declared registration policy, obtains a candidate
score, calibrates candidate probabilities on inner held-out partitions of the
outer-training trials, and reports

\[
\mathrm{gaze\_info\_bits}
=
\log_2\frac{p(\mathrm{correct}\mid\mathrm{gaze})}
{p(\mathrm{correct})}.
\]

Zero bits means that gaze did not improve on the candidate prior. One bit means
that gaze doubled the probability assigned to the correct item relative to its
prior. With \(K\) equally likely candidates, the maximum is
\(\log_2 K\). Negative values mean that gaze made the correct item less
plausible.

Replay divides its raw duration-grid log likelihood by the number of grid bins
before candidate calibration. The resulting mean log predictive density is
invariant to a proportional change in numerical grid resolution. The residual
calibration temperature has a fixed weak prior centred on one. This prevents a
small, perfectly separated training set from forcing numerically degenerate
zero/one probabilities without tying the prior to an arbitrary grid length. If
a training contrast has fewer than four candidate items, Replay records an
explicit temperature-one fallback instead of pretending that an inner
calibration estimate is available. The raw total HMM likelihood remains in the
alignment object as a diagnostic.

`gaze_info_bits` is the sole primary endpoint. Coverage, spatial error,
chronology, contraction, rank, and alignment stability are nested diagnostics
that explain the score; they are not additional normalized outcomes.

## Choose an engine deliberately

The public workflows make two roles explicit:

- `"replay"` is the directional engine for encoding-to-recall studies. It is a
  directional probability model in which recall is a mixture of locally
  ordered replay, chunk restarts, and candidate-independent background gaze.
- Transport is the symmetric explanatory alignment engine. Use
  `gaze_transport_cv()` when neither path has a privileged direction and
  when one or several separate study presentations may inform each candidate.

There is no universal engine default. The common `gaze_weave_cv()` dispatcher
accepts `engine = "replay"` or `engine = "transport"` and also infers the
engine from `gaze_replay_spec()` or `gaze_transport_spec()`. Transport has a
dedicated entry point because its episode identifiers and declared candidate
priors are part of its estimand rather than optional plotting metadata.

Earlier internal experiments were labelled Transport v1, v2, and v3. Those
were development names, never released APIs. The edge-normalized estimator
formerly labelled v3 is now simply Transport; the older implementations have
been removed. Version labels remain in internal and frozen validation
provenance so the historical evidence stays auditable; they do not identify
selectable methods.

The distinction affects interpretation. Replay's braid contains posterior
correspondence probabilities. Transport's braid contains a regularized optimum;
its diffuseness is partly controlled by numerical regularization and is not a
posterior uncertainty statement.

## A cross-fitted contraction example

The following paths differ by a known source-to-reference contraction and
translation. Eight items keep the example compact while permitting an inner
calibration split; scientific analyses should use more training items and
predeclare the candidate pool.

```{r data}
make_path <- function(item) {
  anchor <- c(item * 2.5, (item %% 2) * 2.8)
  xy <- rbind(
    anchor,
    anchor + c(0.8, 1.1),
    anchor + c(1.7, -0.2),
    anchor + c(2.1, 0.7)
  )
  fixation_group(
    x = xy[, 1], y = xy[, 2],
    duration = c(1, 2, 1, 1), onset = c(0, 1, 3, 4)
  )
}

encoding_paths <- lapply(1:8, make_path)
recall_paths <- lapply(encoding_paths, function(path) {
  xy <- sweep(cbind(path$x, path$y), 2, c(0.4, -0.25), FUN = "-") / 0.75
  fixation_group(
    x = xy[, 1], y = xy[, 2],
    duration = path$duration * 1.5,
    onset = path$onset * 1.5
  )
})

encoding <- tibble(
  participant = "p1", image_id = 1:8, fixgroup = encoding_paths
)
recall <- tibble(
  participant = "p1", image_id = 1:8, fixgroup = recall_paths
)
```

The warp is learned only from training items and then applied identically to
every candidate for a held-out path. Pair-specific free registration is not the
default because it can rescue wrong templates.

```{r fit}
replay_spec <- gaze_replay_spec(
  grid_size = 24,
  transition_grid = list(
    background = c(0.02, 0.08),
    restart = c(0.02, 0.08),
    advance = c(0.2, 0.4),
    background_stay = 0.9
  ),
  warp = gaze_warp_contraction(
    center = c(0, 0), translation = TRUE, fit_by = "participant"
  )
)

fit <- gaze_weave_cv(
  ref_tab = encoding,
  source_tab = recall,
  match_on = c("participant", "image_id"),
  contrast_on = "participant",
  split_on = c("participant", "image_id"),
  n_folds = 2,
  seed = 4,
  engine = "replay",
  spec = replay_spec
)

broom::tidy(fit)
```

`tidy()` returns only identifiers and `gaze_info_bits`. Candidate probability,
log loss, rank, convergence, and fold columns are available through
`broom::tidy(fit, diagnostics = TRUE)`. The complete object retains candidate
tables and the true-template alignment for audit and plotting.

## Read the gaze braid

```{r overlay, fig.cap="Encoding, raw recall, registered recall, and posterior replay destinations."}
autoplot(
  fit,
  key = list(participant = "p1", image_id = 1),
  type = "overlay"
)
```

```{r braid, fig.cap="Replay posterior correspondence through normalized gaze time."}
autoplot(
  fit,
  key = list(participant = "p1", image_id = 1),
  type = "braid"
)
```

Parallel ribbons support local order preservation. Crossings can reflect chunk
restarts or reordering. Faint recall time points support the background state.
The diagnostic panel reports posterior replay coverage, background coverage,
restart rate, and spatial RMSE. Those quantities explain the fitted replay but
do not add up to `gaze_info_bits`, which is a contrast across candidate
templates.

## Fit symmetric Transport

Transport replaces path-length-diluted chronology with bounded directed
local-edge disagreement. It integrates matched coverage under a fixed prior,
separates source and reference selection from correspondence, and applies one
candidate-invariant warp to every candidate in a held-out set. Its coupling is
an optimized alignment correspondence. It is never a posterior replay
probability.

The example below turns each encoding path into two separate study
presentations. The presentations are not concatenated: each candidate receives
the mean of its two calibrated episode likelihoods. The one-presentation case
uses exactly the same formula with weight one.

```{r transport-data}
study <- expand.grid(
  participant = "p1",
  image_id = 1:8,
  presentation = 1:2,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
) %>%
  as_tibble()
study$fixgroup <- lapply(seq_len(nrow(study)), function(row) {
  path <- encoding_paths[[study$image_id[[row]]]]
  presentation <- study$presentation[[row]]
  fixation_group(
    x = path[["x"]] + 0.03 * (presentation - 1),
    y = path[["y"]] - 0.02 * (presentation - 1),
    duration = path[["duration"]],
    onset = path[["onset"]]
  )
})

transport_spec <- gaze_transport_spec(
  spatial = gaze_gaussian_mixture(c(0.5, 1), unit = "native"),
  warp = gaze_warp_contraction(
    center = c(0, 0), translation = TRUE, fit_by = "participant"
  ),
  # Lightweight deterministic controls for this executable vignette.
  coverage_nodes = 2,
  entropy_schedule = 0.03,
  maxit = 40,
  tolerance = 1e-3,
  projection_maxit = 300,
  projection_tolerance = 1e-7,
  backend = "optimized",
  reliability = "none"
)
```

The call declares the candidate stratum, outer split, episode identifier, and
candidate prior. `priorvar = NULL` means a uniform design; a positive reference
column supplies a nonuniform design prior. The compact solver controls above
keep this vignette quick; a scientific protocol should freeze its coverage,
solver, calibration, and reliability settings before outcomes are examined.

```{r transport-fit}
transport_fit <- gaze_transport_cv(
  ref_tab = study,
  source_tab = recall,
  match_on = c("participant", "image_id"),
  contrast_on = "participant",
  split_on = c("participant", "image_id"),
  episode_on = "presentation",
  priorvar = NULL,
  n_folds = 2,
  seed = 20260822,
  spec = transport_spec
)

broom::tidy(transport_fit)
```

Default tidying exposes identifiers and `gaze_info_bits` only. Ask for the
audit view when you need the explanatory fields:

```{r transport-audit}
transport_audit <- gaze_transport_result(
  transport_fit,
  key = list(participant = "p1", image_id = 1)
)

transport_audit$diagnostics[c(
  "matched_coverage", "spatial_rmse", "spatial_rmse_unit",
  "local_order_preservation", "contraction_scale", "template_rank",
  "episode_count", "episode_normalization", "solver_stability"
)]
```

Spatial RMSE is reported in the declared screen or spatial coordinate unit; a
`"native"` specification is labelled `native_coordinate_units`. The nested
episode table retains every equal weight, pair score, alignment diagnostic,
backend, and convergence result.

The evidence panel displays the one inferential score. The other panels show a
named episode so their provenance is visible; choosing another episode changes
only the explanatory alignment panel, never the equal-prior score.

```{r transport-evidence, fig.cap="The one-score Transport evidence ledger."}
autoplot(
  transport_fit,
  key = list(participant = "p1", image_id = 1),
  type = "evidence"
)
```

```{r transport-braid, fig.cap="An optimized Transport correspondence, not a posterior probability."}
autoplot(
  transport_fit,
  key = list(participant = "p1", image_id = 1),
  type = "braid",
  episode = transport_audit$diagnostics$episodes$episode[[1]]
)
```

Use `type = "overlay"`, `"raw_registered"`, or `"diagnostics"` for the
registered spatial overlay, the raw-versus-registered path comparison, or the
bounded diagnostic ledger. The public deterministic path fixtures and their
MD5 manifest live under
`inst/validation/gaze-weave-transport-v3-golden` in the source and installed
package; the executable vignette above uses only generated paths. The `v3` in
those frozen paths identifies the development court, not a selectable method.

## What the Transport courts support

The public frozen court passed representation, null calibration, chronology,
coverage-refinement, registration, convergence, and perturbation gates.
Reversal sensitivity stayed stable across the declared path-length range;
12-to-24-node coverage refinement changed pair energy by at most `9.65e-4`
and candidate information by `4.60e-4` bits. Optimized and reference scores
agree within `1e-8` on the differential court. On the frozen local runner, the
15-by-15 optimized pair took 8 ms versus 47 ms for the R reference, and the
scope-matched 315-candidate fold took 13.955 seconds versus 830.849 seconds for
the GW-14 implementation. These are local macOS arm64 measurements, not
cross-platform guarantees.

The response-blind full-cohort study court also passed. P1-to-P4 Transport
information was `0.28464` bits (crossed 95% interval `0.18206` to `0.38573`),
and the intact-minus-reversal/shuffle contrast was `0.05237` bits (`0.00720`
to `0.09536`). Transport was effectively tied with the earlier development
comparator then labelled v2, registered density, density sigma 80, and Replay
on the common panel.

The exact frozen 0--3000 ms retrieval court did not support a retrieval claim.
Transport had the largest response-blind point estimate, `0.02719` bits,
but its crossed interval was `-0.00305` to `0.05618` and principal comparator
contrasts included zero. Adding Transport information to the held-out old-item
response model changed predictive information by `-0.01508` bits (`-0.04322`
to `0.00624`). No algorithm, time window, calibration, subgroup, or saliency
form was changed after behavior was opened.

The resulting product role is bounded: Transport is the symmetric
explanatory alignment engine for order- and partial-replay questions. It is not
a default retrieval predictor and is not known to outperform Replay,
registered density, or MultiMatch generally. The complete public
protocols, aggregate reports, and freeze receipts are installed under
`inst/validation`; trial-linked private outputs are excluded from Git and
package artifacts.

## Invariance contract

The recommended warp is participant/session contraction plus a calibration
translation learned out of fold. Affine and pair-specific warps are not default
nuisance corrections. Coordinate units, candidate pool, design prior, valid
episode intersection, and episode count are part of the scientific estimand
and must be reported.

Transport is invariant to candidate order, batch size, harmless time
dilation, and fixation splitting that preserves the mass/edge representation
within the frozen tolerances. It is deliberately sensitive to local-order
reversal and coherent partial coverage. Results can fail or become
uninterpretable when folds leak participant/item cells, candidate pools differ
across conditions, coordinate units are undeclared, non-identity registration
is fit on the scored pair, too few inner rows force identity calibration, or
the solver does not converge. Treat those conditions as design or diagnostic
failures, not as evidence of absent replay. New analyses should use
`gaze_transport_*` for symmetric alignment or `gaze_replay_*` for the
directional model and should treat `gaze_info_bits` as the primary endpoint.
