---
title: Latent Transforms for Template Similarity
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Latent Transforms for Template Similarity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: violet
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-violet');});</script>

---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

```{r load-packages}
library(eyesim)
library(dplyr)
library(ggplot2)
```

## Why latent transforms?

`template_similarity()` compares density maps in raw screen space. But when
encoding and retrieval differ by a linear transform — different screen sizes,
calibration drift between sessions, or systematic participant differences —
raw correlation can underestimate true reinstatement.

Latent transforms address this by mapping both sides into a shared space
before computing similarity. You add them to `template_similarity()` with
two arguments; the core API stays unchanged.

## A concrete example

Suppose two devices record eye-movements on the same stimuli, but one device
scales coordinates differently. Let's simulate this:

```{r build-data}
set.seed(1)

make_density <- function(vec) {
  fg <- fixation_group(
    x = vec[1:2] * 50 + 50,
    y = vec[3:4] * 50 + 50,
    onset = c(0, 100),
    duration = c(100, 100)
  )
  eye_density(fg, sigma = 30, xbounds = c(0, 100), ybounds = c(0, 100))
}

base_vecs <- replicate(8, rnorm(4), simplify = FALSE)
```

Now create a "source" set that is a scaled version of the reference — as if
recorded on a differently-calibrated device:

```{r apply-scaling}
scale_factors <- c(2, 0.6, 1.7, 0.8)
source_vecs <- lapply(base_vecs, function(v) v * scale_factors)

ref_tab <- tibble(id = seq_along(base_vecs),
                  density = lapply(base_vecs, make_density))
source_tab <- tibble(id = seq_along(source_vecs),
                     density = lapply(source_vecs, make_density))
```

## Comparing raw vs. CORAL-transformed similarity

Without any transform, the scaling distortion reduces similarity:

```{r raw-similarity}
raw <- template_similarity(ref_tab, source_tab,
                           match_on = "id",
                           permutations = 0,
                           method = "cosine")
cat("Raw mean similarity:", round(mean(raw$eye_sim), 3), "\n")
```

CORAL (CORrelation ALignment) whitens the source domain and re-colors it with
the reference covariance, correcting for the scaling difference:

```{r coral-similarity}
coral_res <- template_similarity(
  ref_tab, source_tab,
  match_on = "id",
  permutations = 0,
  method = "cosine",
  similarity_transform = coral_transform,
  similarity_transform_args = list(comps = 4, shrink = 1e-6)
)
cat("CORAL mean similarity:", round(mean(coral_res$eye_sim), 3), "\n")
```

```{r plot-comparison, echo = FALSE, fig.cap = "Paired comparison of raw vs. CORAL-transformed similarity for each item. CORAL recovers higher similarity by correcting the scaling distortion."}
comparison <- tibble(
  id = raw$id,
  Raw = raw$eye_sim,
  CORAL = coral_res$eye_sim
) %>%
  tidyr::pivot_longer(-id, names_to = "method", values_to = "similarity")

ggplot(comparison, aes(x = factor(id), y = similarity, fill = method)) +
  geom_col(position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = c(Raw = "grey60", CORAL = "steelblue")) +
  labs(x = "Item", y = "Cosine Similarity", fill = "Method") +
  theme_minimal()
```

CORAL recovers substantially higher similarity by correcting the covariance
mismatch between the two "devices."

## Available transforms

| Transform | Supervised? | Best for |
|:----------|:------------|:---------|
| `latent_pca_transform()` | No | Dimensionality reduction, mild noise smoothing |
| `coral_transform()` | No | Device/participant shifts (covariance-level differences) |
| `cca_transform()` | Yes (needs matched pairs) | Item-level alignment when pairings are reliable |

All are passed to `template_similarity()` via the `similarity_transform`
argument:

```r
template_similarity(
  ref_tab, source_tab,
  match_on = "id",
  similarity_transform = cca_transform,
  similarity_transform_args = list(comps = 10, shrink = 0.01)
)
```

## Choosing a transform

- **PCA** is the safest default — dimensionality reduction with no assumptions
  about domain differences. Use when you want noise smoothing without domain
  adaptation.
- **CORAL** is unsupervised and assumes linear, covariance-level differences
  between domains. Good for calibration drift or different screen sizes.
- **CCA** is supervised and leverages matched pairs to find shared latent axes.
  Set `comps` modestly (5–15) and add `shrink` to stabilize small-N fits.

After the transform, `template_similarity()` still uses your chosen `method`
(cosine, pearson, fisherz, etc.) on the transformed vectors.

## Notes and limitations

- Multiscale densities are supported only when all scales share the same grid
  size; otherwise latent transforms will error.
- Regularization (`shrink`) and component count (`comps`) affect stability on
  small-N data — tune as needed.
- CORAL and CCA are linear; they will not correct nonlinear spatial warps.
  Keep `comps` low to reduce overfitting, especially with few pairs.
- See `vignette("eyesim")` for the core `template_similarity()` workflow
  without transforms.
