---
title: "Latent Transforms for Template Similarity"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Latent Transforms for Template Similarity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(eyesim)
library(dplyr)
```

## Why latent transforms?

`template_similarity()` compares densities in screen space. When encoding and retrieval differ by a linear transform (e.g., scale/translation) or domain shift (participant/device), raw cosine/correlation can underestimate true reinstatement. Latent transforms insert an optional pre-step that maps both sides into a shared space before similarity. This keeps the core API unchanged while allowing domain adaptation.

### Available transforms

- `latent_pca_transform()`: PCA projection to a fixed number of components. Use when you want dimensionality reduction and mild noise smoothing.
- `coral_transform()`: CORAL (CORrelation ALignment) whitens the source domain and re-colors it with the reference covariance. Unsupervised; good for device/participant shifts when differences are mostly second-moment changes.
- `cca_transform()`: Regularized CCA on matched pairs in PCA space. Supervised; emphasizes components that covary across domains. Use when pairings are reliable and you want item-level alignment.

All are invoked via `template_similarity(..., similarity_transform = <fn>, similarity_transform_args = list(...))`.

## Minimal example

Here we build two tiny density tables where the source is a scaled version of the reference. CORAL should recover higher similarity than the raw comparison.

```{r}
make_density_vec <- function(vec) {
  structure(list(
    z = matrix(vec, nrow = 2, ncol = 2, byrow = TRUE),
    x = 1:2,
    y = 1:2,
    sigma = 50
  ), class = c("density", "eye_density"))
}

set.seed(1)
base_vecs <- replicate(8, rnorm(4, 0, 1), simplify = FALSE)
scale_mat <- diag(c(2, 0.6, 1.7, 0.8))
source_vecs <- lapply(base_vecs, function(v) as.vector(scale_mat %*% v))

ref_tab <- tibble(id = seq_along(base_vecs), density = lapply(base_vecs, make_density_vec))
source_tab <- tibble(id = seq_along(source_vecs), density = lapply(source_vecs, make_density_vec))

raw <- template_similarity(ref_tab, source_tab, match_on = "id", permutations = 0, method = "cosine")
coral <- template_similarity(
  ref_tab, source_tab,
  match_on = "id", permutations = 0, method = "cosine",
  similarity_transform = coral_transform,
  similarity_transform_args = list(comps = 4, shrink = 1e-6)
)

mean(raw$eye_sim)
mean(coral$eye_sim)
```

## Choosing a transform

- **PCA**: default choice for dimensionality reduction; low risk, no pairing needed.
- **CORAL**: unsupervised domain shift; assumes linear, covariance-level differences (e.g., calibration drift, different screen sizes).
- **CCA**: supervised; leverages matched pairs to find shared latent axes. Set `comps` modestly (5–15) and add `shrink` to stabilize small-N fits.

After the transform, `template_similarity` still uses your chosen `method` (cosine, pearson, fisherz, etc.) on the transformed vectors.

## Notes and limitations

- Multiscale densities are supported only when all scales share the same grid size; otherwise latent transforms will error.
- These transforms operate on flattened densities; downstream similarity still uses the chosen `method`.
- Regularization (`shrink`) and component count (`comps`) can affect stability on small-N data—tune as needed.
- CORAL and CCA are linear; they will not correct nonlinear spatial warps. Keep `comps` low to reduce overfitting, especially with few pairs.
