---
title: Comparing Scanpaths with MultiMatch
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Comparing Scanpaths with MultiMatch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-violet');});</script>

---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE))
  ggplot2::theme_set(albersdown::theme_albers(params$family))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

```{r load-packages}
library(eyesim)
library(dplyr)
library(ggplot2)
library(patchwork)
library(ggridges)
```

A single similarity number can hide important differences between scanpaths.
Two people might fixate the same locations but in a completely different order,
or follow the same trajectory but at different speeds. MultiMatch
(Dewhurst et al., 2012) addresses this by evaluating scanpath similarity
across five distinct dimensions:

| Dimension | What it captures |
|:----------|:-----------------|
| **vector** | Overall shape similarity of saccade vectors |
| **direction** | Angular similarity of saccade directions |
| **length** | Similarity of saccade amplitudes |
| **position** | Spatial proximity of corresponding fixations |
| **duration** | Similarity of fixation durations |

The `eyesim` implementation adds a sixth metric, **position_emd**, based on
the earth mover's distance between fixation locations. This captures
distributional similarity that is not order-dependent.

## How do you compare two scanpaths?

Start by creating two `fixation_group` objects, convert them to `scanpath`
objects, and call `multi_match()`:

```{r simulate-linear}
set.seed(1)

simulate_linear <- function(n) {
  fixation_group(
    x = cumsum(rnorm(n, mean = 50, sd = 10)),
    y = cumsum(rnorm(n, mean = 50, sd = 10)),
    duration = rpois(n, lambda = 200),
    onset = 1:n
  )
}

fg1 <- simulate_linear(10)
fg2 <- simulate_linear(12)
```

```{r plot-linear, echo = FALSE, fig.cap = "Two scanpaths generated from similar linear processes."}
plot(fg1, legend = FALSE) + plot(fg2, legend = FALSE)
```

```{r compare-linear}
sp1 <- scanpath(fg1)
sp2 <- scanpath(fg2)
multi_match(sp1, sp2, screensize = c(500, 500))
```

All metrics are high, consistent with the known similarity of the two
scanpaths.

## What happens when scanpaths differ structurally?

Let's compare a linear scanpath against a zigzag pattern:

```{r compare-zigzag}
fg_zigzag <- fixation_group(
  x = cumsum(rep(50, 10)),
  y = cumsum(c(50, -50, 50, -50, 50, -50, 50, -50, 50, -50)),
  duration = rpois(10, lambda = 200),
  onset = 1:10
)

sp_linear <- scanpath(simulate_linear(10))
sp_zigzag <- scanpath(fg_zigzag)
multi_match(sp_linear, sp_zigzag, screensize = c(500, 500))
```

Notice how the **direction** metric drops substantially — the zigzag and
linear paths follow very different angular trajectories.

## What do the metrics look like for random scanpaths?

To build intuition about each metric's range and distribution, let's simulate
500 pairs of random scanpaths and plot the results:

```{r simulate-random, include = FALSE}
set.seed(42)
fgs <- replicate(500, {
  fg1 <- fixation_group(runif(10) * 500, runif(10) * 500,
                        round(runif(10) * 10) + 1, 1:10)
  fg2 <- fixation_group(runif(10) * 500, runif(10) * 500,
                        round(runif(10) * 10) + 1, 1:10)
  list(fg1 = fg1, fg2 = fg2)
}, simplify = FALSE)

ssdf <- lapply(fgs, function(x) {
  multi_match(scanpath(x$fg1), scanpath(x$fg2), c(500, 500))
}) %>% bind_rows()
```

```{r plot-distributions, fig.height = 4.5, echo = FALSE, fig.cap = "Distribution of each MultiMatch metric across 500 random scanpath pairs, on a common 0 to 1 scale. Vertical lines mark the medians."}
metric_labels <- c(mm_vector = "Vector", mm_direction = "Direction",
                   mm_length = "Length", mm_position = "Position",
                   mm_duration = "Duration", mm_position_emd = "Position (EMD)")
long <- tidyr::pivot_longer(ssdf[names(metric_labels)], dplyr::everything(),
                            names_to = "metric", values_to = "similarity")
long$metric <- factor(metric_labels[long$metric], levels = rev(metric_labels))
medians <- tapply(long$similarity, long$metric, stats::median)

ggplot(long, aes(x = similarity, y = metric)) +
  geom_density_ridges(scale = 0.9, rel_min_height = 0.01, bandwidth = 0.03,
                      fill = eyesim_colours("correspondence"), alpha = 0.35,
                      colour = eyesim_colours("correspondence"),
                      quantile_lines = TRUE, quantiles = 2) +
  scale_x_continuous(limits = c(0, 1), expand = expansion(mult = 0.01)) +
  labs(x = "Similarity (1 = identical)", y = NULL) +
  theme_eyesim()
```

None of the metrics sits near zero for unrelated scanpaths. Median
similarity for random pairs ranges from
`r round(min(medians), 2)` (`r names(which.min(medians))`) to
`r round(max(medians), 2)` (`r names(which.max(medians))`), so a
single similarity value has to be read against this baseline, not against
zero.

## What do extreme matches look like?

Examining the highest- and lowest-scoring pairs for each metric gives concrete
insight into what each dimension captures:

```{r compute-extremes, include = FALSE}
metric_names <- c(vector = "Vector", direction = "Direction", length = "Length",
                  position = "Position", duration = "Duration",
                  position_emd = "Position (EMD)")

# One row per fixation: metric, which extreme (highest or lowest scoring
# pair) and which scanpath of the pair.
extreme_fixations <- do.call(rbind, lapply(names(metric_names), function(m) {
  col <- paste0("mm_", m)
  do.call(rbind, lapply(c(Highest = "which.max", Lowest = "which.min"), function(f) {
    i <- match.fun(f)(ssdf[[col]])
    extreme <- if (f == "which.max") "Highest" else "Lowest"
    do.call(rbind, lapply(1:2, function(k) {
      fg <- fgs[[i]][[k]]
      data.frame(
        metric = sprintf("%s\n%s %.2f / %s %.2f", metric_names[[m]],
                         "high", max(ssdf[[col]]), "low", min(ssdf[[col]])),
        panel = paste(extreme, "pair, scanpath", c("A", "B")[k]),
        x = fg$x, y = fg$y, onset = fg$onset, duration = fg$duration,
        order = rank(fg$onset, ties.method = "first")
      )
    }))
  }))
}))
extreme_fixations$metric <- factor(extreme_fixations$metric,
                                   levels = unique(extreme_fixations$metric))
```

```{r plot-extremes, fig.height = 11, echo = FALSE, fig.cap = "The highest- and lowest-scoring random pairs for each MultiMatch metric (rows). Each pair is shown as its two scanpaths side by side; colour is fixation order and point area is duration. Row labels give the high and low scores."}
extreme_fixations <- extreme_fixations[order(extreme_fixations$order), ]
ggplot(extreme_fixations, aes(x, y)) +
  geom_path(aes(colour = order), linewidth = 0.4, alpha = 0.8) +
  geom_point(aes(fill = order, size = duration), shape = 21,
             colour = "white", stroke = 0.3) +
  geom_point(data = subset(extreme_fixations, order == 1), shape = 21,
             size = 4, fill = NA, colour = eyesim_colours("ink"), stroke = 0.5) +
  facet_grid(metric ~ panel, switch = "y") +
  scale_colour_eyesim_time(aesthetics = c("colour", "fill"), name = "Fixation\norder",
                           breaks = c(1, 4, 7, 10)) +
  scale_size_area(max_size = 2.8, name = "Duration",
                  guide = guide_legend(override.aes = list(fill = "grey45"))) +
  coord_fixed(xlim = c(0, 500), ylim = c(0, 500), expand = TRUE) +
  labs(caption = "The ringed point is each scanpath's first fixation.") +
  theme_eyesim_spatial(base_size = 9) +
  theme(strip.text.y.left = element_text(angle = 0, hjust = 1),
        strip.placement = "outside", panel.spacing = unit(4, "pt"),
        legend.position = "bottom")
```

The highest-scoring pairs are not visibly similar scanpaths: with ten random
fixations each, a pair can score highly on one dimension (for example,
similar saccade lengths) while differing everywhere else. This is why the
five dimensions are reported separately rather than averaged.

## How do transformations affect the metrics?

Understanding how each metric responds to spatial and temporal transformations
helps you choose the right one for your research question.

### Identity (perfect match)

```{r transform-identity}
set.seed(7)
fg <- fixation_group(runif(10) * 500, runif(10) * 500,
                     round(runif(10) * 10) + 1, 1:10)
multi_match(scanpath(fg), scanpath(fg), c(500, 500))
```

All metrics are 1.0 — a scanpath is perfectly similar to itself.

### Scaling

Scaling the coordinates by 0.5 preserves **direction** (relative
angles are unchanged) but reduces **position** (absolute locations differ):

```{r transform-scale}
fg_scaled <- fg
fg_scaled$x <- fg_scaled$x * 0.5
fg_scaled$y <- fg_scaled$y * 0.5
```

```{r plot-scale, echo = FALSE, fig.cap = "Original (left) vs. scaled (right) scanpath."}
plot(fg, xlim = c(0, 500), ylim = c(0, 500), legend = FALSE) +
  plot(fg_scaled, xlim = c(0, 500), ylim = c(0, 500), legend = FALSE)
```

```{r compare-scale}
multi_match(scanpath(fg), scanpath(fg_scaled), c(500, 500))
```

As expected, **direction** stays at 1.0, but **position** drops.

### Shuffling fixation order

Keeping the same fixation locations but scrambling their order preserves
**position** (and the EMD) while disrupting **direction**:

```{r transform-shuffle}
set.seed(3)
fg_shuffled <- fg
ord <- sample(nrow(fg_shuffled))
fg_shuffled$x <- fg_shuffled$x[ord]
fg_shuffled$y <- fg_shuffled$y[ord]
fg_shuffled$duration <- fg_shuffled$duration[ord]
```

```{r plot-shuffle, echo = FALSE, fig.cap = "Original (left) vs. order-shuffled (right) scanpath. Same locations, different sequence."}
plot(fg, xlim = c(0, 500), ylim = c(0, 500), legend = FALSE) +
  plot(fg_shuffled, xlim = c(0, 500), ylim = c(0, 500), legend = FALSE)
```

```{r compare-shuffle}
multi_match(scanpath(fg), scanpath(fg_shuffled), c(500, 500))
```

The **position_emd** remains high (same spatial distribution), while
**direction** drops (the trajectory has changed).

## References

Dewhurst, R., Nyström, M., Jarodzka, H., Foulsham, T., Johansson, R., &
Holmqvist, K. (2012). It depends on how you look at it: Scanpath comparison in
multiple dimensions with MultiMatch, a vector-based approach. *Behavior
Research Methods*, 44, 1079–1100.
