---
title: "Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Eye-movement similarity analysis

The main goal of `eyesim` is to provide tools for computing measures of similarity between eye-movement fixation data collected over a series of trials embedded in an experimental design. A major focus of the library is to offer easy ways to compare fixation patterns between two experimental states, for example, between perceiving an image and remembering that same image. These kinds of analyses are useful in assessing and quantifying so-called "eye-movement reinstatement" in studies of memory. Below we describe some basic aspects of the library that should allow one to get started with it.


## A Basic unit: the fixation group

A fixation group is a set of eye-movement fixations that comprise a meaningful unit in a study, for example, a trial, a condition, a participants, a time window, etc. Every fixation group contains a set of xy coordinates and corresponding vectors indicating the `onset` (when did the fixation start?) and the `duration` (how long did the fixation last?) of the sequence of fixations. Below we create a fixation group from a set of 3 coordinates occurring at at times 0, 10, and 60.

After creating a `fixation_group` object, we then plot it to visualize the location of the three fixations. The size of each point is scaled by duration and the color of each point is mapped to the onset time, with yellow colors being early-ocurring and red colors being late-occurring in this group.


```{r, message=FALSE, warning=FALSE, fig.width=8}
library(eyesim)
library(patchwork)
library(dplyr)
library(ggplot2)
fg <- fixation_group(x=c(-100, 0, 100), y=c(0, 100, 0), onset=c(0,10,60), duration=c(10,50,100))
plot(fg)

```

Now we create larger group consisting of 25 randomly generated eye-movements.

```{r}

## generate random fixation coordinates
cds <- do.call(rbind, lapply(1:25, function(i) {
  data.frame(x=runif(1)*100, y=runif(1)*100)
}))

## generate random increasing onsets
onset <- cumsum(runif(25)*100)

## construct a "fixation_group" object
fg <- fixation_group(x=cds[,1], y=cds[,2], onset=onset, duration=c(diff(onset),25))
plot(fg)

```

That's a pretty busy display, but captures the sequence of eye-movements. We can also plot various kinds of "density" maps which show where the fixations were most likely to occur.

Below we show three different ways of plotting fixation density: a contour plot, a raster plot, and a filled contour plot.


```{r, fig.width=8}

p1 <- plot(fg, typ="contour", xlim=c(-10,110), ylim=c(-10,110), bandwidth=35)
p2 <- plot(fg, typ="raster", xlim=c(-10,110), ylim=c(-10,110), bandwidth=35)
p3 <- plot(fg, typ="filled_contour", xlim=c(-10,110), ylim=c(-10,110), bandwidth=35)

p1+p2+p3
```


We can also vary the bandwidth of the two-dimensional density estimation procedure to visualize the fixations at different levels of smoothness.


```{r, fig.width=8}

p1 <- plot(fg, typ="filled_contour", xlim=c(-10,110), ylim=c(-10,110), bandwidth=20)
p2 <- plot(fg, typ="filled_contour", xlim=c(-10,110), ylim=c(-10,110), bandwidth=60)
p3 <- plot(fg, typ="filled_contour", xlim=c(-10,110), ylim=c(-10,110), bandwidth=100)

p1+p2+p3

```

## Computing similarity between fixation groups

Suppose we have two fixation groups, `fg1` and `fg2`, how do we compare these coordinate sets? The `eyesim` package provides methods for computing similarities between spatial density maps. Below, we generate two eye-movement patterns, one of which is a perturbed version of the other. Then we compute a series of similarity metrics on the two patterns.

```{r, fig.width=7}


cds <- do.call(rbind, lapply(1:25, function(i) {
  data.frame(x=runif(1)*100, y=runif(1)*100)
}))

cds2 <- do.call(rbind, lapply(1:25, function(i) {
  if (i %% 2 == 0) {
    data.frame(x=runif(1)*100, y=runif(1)*100)
  } else {
    data.frame(x=cds[i,1], y=cds[i,2])
  }
}))

onset <- cumsum(runif(25)*100)
fg1 <- fixation_group(x=cds[,1], y=cds[,2], onset=onset, duration=c(diff(onset),25))
fg2 <- fixation_group(x=cds2[,1], y=cds2[,2], onset=onset, duration=c(diff(onset),25))

p1 <- plot(fg1)
p2 <- plot(fg2)
p1+p2
```

To compute the similarity between any two `fixation_group`s we use the `similarity` generic function. First we convert the `fixation_group`s into `eye_density` objects and then compute their similarity. The default metric for comparing two density maps is the Pearson correlation coefficient.

```{r}

ed1 <- eye_density(fg1, sigma=50, xbounds=c(0,100), ybounds=c(0,100))
ed2 <- eye_density(fg2, sigma=50, xbounds=c(0,100), ybounds=c(0,100))
ed3 <- eye_density(fg2, sigma=50, xbounds=c(0,100), ybounds=c(0,100), duration_weighted = TRUE)

s1 <- similarity(ed1,ed2)
s1

```

We can compute other similarity measures as well. Below we compute similarity using the Pearson correlation, Spearman correlation, the Fisher-transformed Pearson correlation, the cosine similarity, the "l1" similarity based on the 1-norm, the [Jaccard similarity](https://en.wikipedia.org/wiki/Jaccard_index) (from `proxy` package), and the [distance covariance](https://en.wikipedia.org/wiki/Distance_correlation) (`dcov` from the `energy` package).

```{r}

methods=c("pearson", "spearman", "fisherz", "cosine", "l1", "jaccard", "dcov")
for (meth in methods) {
  s1 <- similarity(ed1,ed2, method=meth)
  print(paste(meth, ":", s1))
}

```

## Computing similarity between set of fixation groups in an experiment

Suppose we have a memory experiment in which images are presented during an "encoding" block and also during a retrieval/recognition block. Some images are repeated (or cued in some way) and subjects asked to recognize (or recall) the images. In this situation, we might want to compute the pairwise similarity between encoding and retrieval trials when the image was repeated. We might also want to control for non-specific eye-movement similarity between any two arbitrary encoding and retrieval trials.

Here we will simulate data from an experiment in which 50 images are shown during an encoding block and then the same 50 images are shown (or cued) during a retrieval block. We will then compute the eye-movement similarity between the encoding and retrieval trials corresponding to the same imege and the average similarity between a set of non-corresponding encoding and retrieval trials. 

We will generate data for three participants, each of which has 50 encoding and 50 retrieval trials. We will use the `eye_table` data structure, which is an extension of `data.frame` to hold the data. 

The the code below, the function `gen_fixations` generates a number of fixations (between 1 and 10) that are randomly distributed in a 100 by 100 coordinate frame. Although the fixations for every trial are randomly selected, we assign an experimental condition ("encoding", "retrieval") and subject id ("s1", "s2", "s3") to each set of generated coordinates.


```{r}

gen_fixations <- function(imname, phase, trial, participant) {
  ## generate some number of fixation between 1 and 10
  nfix <- ceiling(runif(1) * 10)
  cds <- do.call(rbind, lapply(1:nfix, function(i) {
    data.frame(x=runif(1)*100, y=runif(1)*100)
  }))
  
  onset <- cumsum(runif(nfix)*100)
  df1 <- data.frame(x=cds[,1], y=cds[,2], onset=onset, 
                    duration=c(diff(onset),100), image=imname, 
                    phase=phase, trial=trial, participant=participant)

}

df1 <- lapply(c("s1", "s2", "s3"), function(snum) {
  lapply(c("encoding", "retrieval"), function(phase) {
    lapply(paste0(1:50), function(trial) {
      gen_fixations(paste0("im", trial), phase, trial, snum)
    }) %>% bind_rows()
  }) %>% bind_rows()
}) %>% bind_rows()



```

Now we are ready to create an `eye_table` data structure which stores the fixations and associates them with the experimental design and a grouping structure. Here we will group the fixations by "image", "participant", and "phase". The will allow sets of fixations to be grouped together in `fixation_group`s so that eye-movement similarity analyses can be carried out. Without grouping variables, there is no way to know how to pool eye-movements together in sets appropriate for kernel density estimation or other analyses.

```{r}

eyetab <- eye_table("x", "y", "duration", "onset", groupvar=c("participant", "phase", "image"), data=df1)

```

```{r, echo=FALSE}

eyetab

```

Next, we will compute the similarity between encoding-retrieval pairs, such that each pair consists of the same image viewed at encoding and retrieval, respectively. In essence, we need to "match" each retrieval trial with its corresponding encoding trial, such that the viewed images in both conditions are the same.

The first step is to compute "density maps" for each fixation_group, defined by the intersection of the participant, image, and phase variables stored in the `eyetab` object.

Below we compute the density maps for all combinations of participant, image, and phase and plot the first four density maps of the resulting set.

```{r, warning=FALSE}

eyedens <- density_by(eyetab, groups=c("phase", "image", "participant"), sigma=100, xbounds=c(0,100), ybounds=c(0,100))

p1 <- plot(eyedens$density[[1]])
p2 <- plot(eyedens$density[[2]])
p3 <- plot(eyedens$density[[3]])
p4 <- plot(eyedens$density[[4]])

(p1 + p2) / (p3+ p4)

```

Below we print the first 10 rows of the resulting data.frame, which contains the `eye_density` objects stored in the `density` variable.

```{r}
eyedens

```

Now that we have the set of `eye_density` maps, we can run a similarity analysis. To do this, we use the `template_similarity` function. We want to compare each "retrieval" density map to its corresponding encoding density map (the "template"). And we want to "match" on the name of the image that was first studied (during "encoding") and later recognized (during "retrieval").

First, we split out the encoding and retrieval trials using the `dplyr::filter` method. Next we call `template_similarity` and indicating that pairs should be matched by the `image` variable. We choose the `fisherz` similarity measure and correct for non-specific eye-movement similarity using 50 permutations in which similarity is computed for non-matching image pairs and subtracted from the similarity score. 

The raw similarity score is returned as `eye_sim` and the permutation-corrected score is returned as `eye_sim_diff`. The similarity score among permuted pairs is also returned as `perm_sim`.

Note on permutations and small-N. The `permutations` argument is an upper bound. If it exceeds the number of available non-matching candidates (determined by `permute_on`), eyesim uses all available candidates (excluding the true match) — i.e., it is exhaustive when possible. For example, with 3 images per participant and `permute_on = participant`, there are only 2 non-matching candidates per trial; setting `permutations = 50` will still use both.

Note on units with `method = "fisherz"`. In this case, `eye_sim`, `perm_sim`, and `eye_sim_diff` are Fisher z scores (atanh of Pearson r). Convert back to correlations with `tanh()` if desired (e.g., `r_diff <- tanh(simres1$eye_sim_diff)`). With `method = "pearson"` or `"spearman"`, values are correlations directly.

Because the eye-movement fixations were generated randomly, we do not expect a non-zero similarity score for the `eye_sim_diff` variable, and we can test that with a one-sample t-test.



```{r, warning=FALSE}
set.seed(1234)
enc_dens <- eyedens %>% filter(phase == "encoding")
ret_dens <- eyedens %>% filter(phase == "retrieval")

simres1 <- template_similarity(enc_dens, ret_dens, match_on="image", method="fisherz", permutations=50)

t.test(simres1$eye_sim_diff)

```

As expected, the t-test is not significant. We can also plot the histograms of similarity scores.

```{r, fig.width=7, warning=FALSE}
par(mfrow=c(1,3))
hist(simres1$eye_sim, main="raw eye movement similarity")
hist(simres1$perm_sim, main="image-permuted eye movement similarity")
hist(simres1$eye_sim_diff, main="corrected eye movement similarity")

```

## Multiscale analysis

Rather than choosing a single bandwidth (sigma) value, we can perform multiscale analysis by providing a vector of sigma values. This approach avoids the need to arbitrarily select a single smoothing parameter and instead captures similarity across multiple spatial scales.

```{r, warning=FALSE}
# Compute multiscale density maps using multiple sigma values
eyedens_multi <- density_by(eyetab, groups=c("phase", "image", "participant"), 
                           sigma=c(25, 50, 100), xbounds=c(0,100), ybounds=c(0,100))

enc_dens_multi <- eyedens_multi %>% filter(phase == "encoding")
ret_dens_multi <- eyedens_multi %>% filter(phase == "retrieval")

# Run template similarity analysis with multiscale data
simres_multi <- template_similarity(enc_dens_multi, ret_dens_multi, 
                                  match_on="image", method="fisherz", permutations=50)

# Compare single-scale vs multiscale results
cat("Single-scale mean similarity:", round(mean(simres1$eye_sim_diff), 4), "\n")
cat("Multiscale mean similarity:", round(mean(simres_multi$eye_sim_diff), 4), "\n")
```

The multiscale approach provides a more robust similarity estimate by averaging across different levels of spatial smoothing.

## Temporal Density Sampling

Sometimes we want to understand how eye-movement patterns relate to a reference density map **over time** during a trial. For example, in memory research, we might ask: "At each moment during retrieval, how much does the current gaze location correspond to where the participant looked during encoding?"

This is different from comparing overall density maps (which `template_similarity` does). Instead, we sample the encoding density at the retrieval fixation locations at each time point, creating a time series of "reinstatement" values.

The `sample_density_time()` function provides this capability. It takes:
- A template table with density maps (e.g., from encoding)
- A source table with fixation groups (e.g., from retrieval)
- Time points at which to sample
- Optional time bins for aggregation

This approach avoids a common problem: if you try to compute separate density maps for short time windows, you often end up with too few fixations (<2) to estimate a density. By using the full encoding density and sampling it over time, we avoid this issue entirely.

```{r, warning=FALSE}
# We already have encoding densities from above
# Now let's sample them at retrieval fixation locations over time

# First, we need an eye_table with fixation groups for retrieval
ret_eyetab <- eyetab %>% filter(phase == "retrieval")

# Create a matching key for both tables
enc_dens <- enc_dens %>% mutate(match_key = paste(participant, image, sep = "_"))
ret_eyetab_matched <- ret_eyetab %>% mutate(match_key = paste(participant, image, sep = "_"))

# Sample encoding density at retrieval fixation locations
# every 50ms from 0-500ms, aggregated into two 250ms bins
temporal_results <- sample_density_time(
 template_tab = enc_dens,
 source_tab = ret_eyetab_matched,
 match_on = "match_key",
 times = seq(0, 500, by = 50),
 time_bins = c(0, 250, 500),
 permutations = 10,
 permute_on = "participant"
)

# View the results
temporal_results %>%
 select(participant, image, bin_1, bin_2, perm_bin_1, perm_bin_2, diff_bin_1, diff_bin_2) %>%
 head()
```

The output includes:
- `sampled`: A list column with the full time series (density value at each time point)
- `bin_1`, `bin_2`, etc.: Aggregated density values for each time bin
- `perm_bin_1`, `perm_bin_2`, etc.: Permuted baseline values (from non-matching templates)
- `diff_bin_1`, `diff_bin_2`, etc.: The difference (observed - permuted)

We can visualize how the sampled density changes over time:

```{r, fig.width=8, fig.height=4, warning=FALSE}
# Extract and plot the time series for a few trials
library(tidyr)

# Get the first 6 trials' time series
time_series_data <- temporal_results %>%
 head(6) %>%
 mutate(trial_id = paste(participant, image, sep = "_")) %>%
 select(trial_id, sampled) %>%
 unnest(sampled)

ggplot(time_series_data, aes(x = time, y = z, color = trial_id)) +
 geom_line(linewidth = 1) +
 labs(x = "Time (ms)", y = "Sampled Density Value",
      title = "Encoding Density Sampled at Retrieval Fixation Locations",
      color = "Trial") +
 theme_minimal() +
 theme(legend.position = "bottom")
```

This temporal analysis reveals when during retrieval participants are looking at locations that were fixated during encoding, which can provide insights into the time course of memory-guided attention.














