---
title: Measuring Similarity Across Repeated Viewings
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Measuring Similarity Across Repeated Viewings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-violet');});</script>

---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE))
  ggplot2::theme_set(albersdown::theme_albers(params$family))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

```{r load-packages}
library(eyesim)
library(dplyr)
library(ggplot2)
```

When participants view the same image multiple times — say, during encoding
and retrieval — do they look at the same locations each time? And is that
consistency specific to the repeated stimulus, or just a general gaze pattern?

`repetitive_similarity()` answers these questions by computing within-stimulus
similarity across experimental conditions and comparing it to cross-stimulus
similarity. Unlike `template_similarity()` which requires explicit
encoding-retrieval pairing, `repetitive_similarity()` examines *all* possible
within-stimulus combinations and summarizes them.

## Setting up the data

Let's simulate a small dataset: 2 participants view 3 images during encoding
and retrieval.

```{r simulate-data}
set.seed(42)

gen_fixations <- function(imname, phase, participant) {
  nfix <- sample(3:10, 1)
  data.frame(
    x = runif(nfix, 0, 100), y = runif(nfix, 0, 100),
    onset = cumsum(runif(nfix, 20, 80)),
    duration = runif(nfix, 80, 300),
    image = imname, phase = phase, participant = participant
  )
}

df <- do.call(rbind, lapply(c("s1", "s2"), function(s) {
  do.call(rbind, lapply(c("encoding", "retrieval"), function(p) {
    do.call(rbind, lapply(paste0("img", 1:3), function(img) {
      gen_fixations(img, p, s)
    }))
  }))
}))

eyetab <- eye_table("x", "y", "duration", "onset",
                    groupvar = c("participant", "phase", "image"),
                    data = df)
```

Compute density maps for each participant-phase-image combination:

```{r compute-densities}
eyedens <- density_by(eyetab,
                      groups = c("phase", "image", "participant"),
                      sigma = 50,
                      xbounds = c(0, 100), ybounds = c(0, 100))
```

## Computing repetitive similarity

`repetitive_similarity()` takes the density table and a condition variable
(here `phase`) that defines the repeated viewings:

```{r compute-repsim}
rep_sim <- repetitive_similarity(eyedens,
                                 condition_var = "phase",
                                 method = "pearson")
rep_sim
```

The result contains two key columns:

- **repsim** — within-stimulus similarity: how consistently participants
  looked at the same locations for the *same* image across conditions
- **othersim** — cross-stimulus similarity: baseline similarity between
  *different* images viewed under different conditions

## Interpreting the results

```{r plot-repsim, fig.cap = "Within-stimulus similarity (repsim). Higher values indicate that participants fixated similar locations when viewing the same image across encoding and retrieval."}
ggplot(rep_sim, aes(x = image, y = repsim)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  labs(x = "Stimulus", y = "Repetitive Similarity",
       title = "Same image across conditions") +
  theme_minimal()
```

```{r plot-othersim, fig.cap = "Cross-stimulus similarity (othersim). This represents the baseline: how similar gaze patterns are when comparing different images across conditions."}
ggplot(rep_sim, aes(x = image, y = othersim)) +
  geom_col(fill = "coral", alpha = 0.7) +
  labs(x = "Stimulus", y = "Other Similarity",
       title = "Different images across conditions") +
  theme_minimal()
```

With random data, both values should be near zero. In a real experiment, you
would expect `repsim` to be higher than `othersim` if participants show
stimulus-specific eye-movement reinstatement. The difference
(`repsim - othersim`) gives you a corrected measure of reinstatement that
controls for general gaze tendencies.

## What's next?

- See `?repetitive_similarity` for additional options, including `pairwise = TRUE`
  to return all pairwise similarities instead of summaries
- `vignette("eyesim")` covers the full `template_similarity()` workflow
  with permutation-based baselines
- `vignette("latent-transforms")` describes domain-adaptation methods that
  can improve similarity estimates across devices or participants
