---
title: "Comparing Scanpaths with MultiMatch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Scanpaths with MultiMatch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',()=>document.body.classList.add('palette-red'));</script>

---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE))
  ggplot2::theme_set(albersdown::theme_albers(params$family))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

```{r load-packages}
library(eyesim)
library(dplyr)
library(ggplot2)
library(patchwork)
library(ggridges)
```

A single similarity number can hide important differences between scanpaths.
Two people might fixate the same locations but in a completely different order,
or follow the same trajectory but at different speeds. MultiMatch
(Dewhurst et al., 2012) addresses this by evaluating scanpath similarity
across five distinct dimensions:

| Dimension | What it captures |
|:----------|:-----------------|
| **vector** | Overall shape similarity of saccade vectors |
| **direction** | Angular similarity of saccade directions |
| **length** | Similarity of saccade amplitudes |
| **position** | Spatial proximity of corresponding fixations |
| **duration** | Similarity of fixation durations |

The `eyesim` implementation adds a sixth metric, **position_emd**, based on
the earth mover's distance between fixation locations. This captures
distributional similarity that is not order-dependent.

## How do you compare two scanpaths?

Start by creating two `fixation_group` objects, convert them to `scanpath`
objects, and call `multi_match()`:

```{r simulate-linear}
set.seed(1)

simulate_linear <- function(n) {
  fixation_group(
    x = cumsum(rnorm(n, mean = 50, sd = 10)),
    y = cumsum(rnorm(n, mean = 50, sd = 10)),
    duration = rpois(n, lambda = 200),
    onset = 1:n
  )
}

fg1 <- simulate_linear(10)
fg2 <- simulate_linear(12)
```

```{r plot-linear, echo = FALSE, fig.cap = "Two scanpaths generated from similar linear processes."}
plot(fg1) + plot(fg2)
```

```{r compare-linear}
sp1 <- scanpath(fg1)
sp2 <- scanpath(fg2)
multi_match(sp1, sp2, screensize = c(500, 500))
```

All metrics are high, consistent with the known similarity of the two
scanpaths.

## What happens when scanpaths differ structurally?

Let's compare a linear scanpath against a zigzag pattern:

```{r compare-zigzag}
fg_zigzag <- fixation_group(
  x = cumsum(rep(50, 10)),
  y = cumsum(c(50, -50, 50, -50, 50, -50, 50, -50, 50, -50)),
  duration = rpois(10, lambda = 200),
  onset = 1:10
)

sp_linear <- scanpath(simulate_linear(10))
sp_zigzag <- scanpath(fg_zigzag)
multi_match(sp_linear, sp_zigzag, screensize = c(500, 500))
```

Notice how the **direction** metric drops substantially — the zigzag and
linear paths follow very different angular trajectories.

## What do the metrics look like for random scanpaths?

To build intuition about each metric's range and distribution, let's simulate
500 pairs of random scanpaths and plot the results:

```{r simulate-random, include = FALSE}
set.seed(42)
fgs <- replicate(500, {
  fg1 <- fixation_group(runif(10) * 500, runif(10) * 500,
                        round(runif(10) * 10) + 1, 1:10)
  fg2 <- fixation_group(runif(10) * 500, runif(10) * 500,
                        round(runif(10) * 10) + 1, 1:10)
  list(fg1 = fg1, fg2 = fg2)
}, simplify = FALSE)

ssdf <- lapply(fgs, function(x) {
  multi_match(scanpath(x$fg1), scanpath(x$fg2), c(500, 500))
}) %>% bind_rows()
```

```{r plot-distributions, fig.height = 10, echo = FALSE, fig.cap = "Distribution of each MultiMatch metric across 500 random scanpath pairs."}
metrics <- c("mm_vector", "mm_direction", "mm_length",
             "mm_position", "mm_duration", "mm_position_emd")
labels <- c("Vector", "Direction", "Length",
            "Position", "Duration", "EMD")
colors <- c("blue", "green", "red", "purple", "orange", "cyan")

plots <- lapply(seq_along(metrics), function(i) {
  ggplot(ssdf, aes(x = .data[[metrics[i]]],
                   y = factor(1, labels = paste(labels[i], "Similarity")))) +
    geom_density_ridges(scale = 3, rel_min_height = 0.01, bandwidth = 0.05,
                        fill = colors[i], alpha = 0.7) +
    labs(x = "Similarity", y = "") +
    theme_ridges() +
    theme(axis.title.y = element_blank())
})

Reduce("/", plots)
```

Each metric has a distinct baseline distribution. Direction similarity, for
instance, clusters near 0.5 for random pairs, while position similarity is
typically lower.

## What do extreme matches look like?

Examining the highest- and lowest-scoring pairs for each metric gives concrete
insight into what each dimension captures:

```{r compute-extremes, include = FALSE}
metric_names <- c("vector", "direction", "length",
                  "position", "duration", "position_emd")

extremes <- lapply(metric_names, function(m) {
  col <- paste0("mm_", m)
  list(high = which.max(ssdf[[col]]), low = which.min(ssdf[[col]]))
})
names(extremes) <- metric_names

plot_fg <- function(fg, title, score) {
  plot(fg) +
    ggtitle(title) +
    annotate("text", x = Inf, y = -Inf,
             label = paste("score:", round(score, 2)),
             hjust = 1.1, vjust = -1.1, size = 2, color = "blue")
}

plots <- lapply(metric_names, function(m) {
  hi <- extremes[[m]]$high
  lo <- extremes[[m]]$low
  col <- paste0("mm_", m)

  list(
    high_plot = plot_fg(fgs[[hi]]$fg1, paste(m, "High(1)"), ssdf[hi, col]) +
                plot_fg(fgs[[hi]]$fg2, paste(m, "High(2)"), ssdf[hi, col]),
    low_plot  = plot_fg(fgs[[lo]]$fg1, paste(m, "Low(1)"), ssdf[lo, col]) +
                plot_fg(fgs[[lo]]$fg2, paste(m, "Low(2)"), ssdf[lo, col])
  )
})
```

```{r plot-high-extremes, fig.height = 11, echo = FALSE, fig.cap = "Highest-scoring pairs for each MultiMatch metric. Each row shows the two scanpaths that scored highest on that dimension."}
wrap_plots(lapply(plots, `[[`, "high_plot"), nrow = 6)
```

```{r plot-low-extremes, fig.height = 11, echo = FALSE, fig.cap = "Lowest-scoring pairs for each metric."}
wrap_plots(lapply(plots, `[[`, "low_plot"), nrow = 6)
```

## How do transformations affect the metrics?

Understanding how each metric responds to spatial and temporal transformations
helps you choose the right one for your research question.

### Identity (perfect match)

```{r transform-identity}
set.seed(7)
fg <- fixation_group(runif(10) * 500, runif(10) * 500,
                     round(runif(10) * 10) + 1, 1:10)
multi_match(scanpath(fg), scanpath(fg), c(500, 500))
```

All metrics are 1.0 — a scanpath is perfectly similar to itself.

### Scaling

Scaling the coordinates by 0.5 preserves **direction** (relative
angles are unchanged) but reduces **position** (absolute locations differ):

```{r transform-scale}
fg_scaled <- fg
fg_scaled$x <- fg_scaled$x * 0.5
fg_scaled$y <- fg_scaled$y * 0.5
```

```{r plot-scale, echo = FALSE, fig.cap = "Original (left) vs. scaled (right) scanpath."}
(plot(fg) + ylim(0, 500)) + (plot(fg_scaled) + ylim(0, 500))
```

```{r compare-scale}
multi_match(scanpath(fg), scanpath(fg_scaled), c(500, 500))
```

As expected, **direction** stays at 1.0, but **position** drops.

### Shuffling fixation order

Keeping the same fixation locations but scrambling their order preserves
**position** (and the EMD) while disrupting **direction**:

```{r transform-shuffle}
set.seed(3)
fg_shuffled <- fg
ord <- sample(nrow(fg_shuffled))
fg_shuffled$x <- fg_shuffled$x[ord]
fg_shuffled$y <- fg_shuffled$y[ord]
fg_shuffled$duration <- fg_shuffled$duration[ord]
```

```{r plot-shuffle, echo = FALSE, fig.cap = "Original (left) vs. order-shuffled (right) scanpath. Same locations, different sequence."}
(plot(fg) + ylim(0, 500)) + (plot(fg_shuffled) + ylim(0, 500))
```

```{r compare-shuffle}
multi_match(scanpath(fg), scanpath(fg_shuffled), c(500, 500))
```

The **position_emd** remains high (same spatial distribution), while
**direction** drops (the trajectory has changed).

## References

Dewhurst, R., Nyström, M., Jarodzka, H., Foulsham, T., Johansson, R., &
Holmqvist, K. (2012). It depends on how you look at it: Scanpath comparison in
multiple dimensions with MultiMatch, a vector-based approach. *Behavior
Research Methods*, 44, 1079–1100.
